---
title: "Practical 4 Answers"
subtitle: "GLMs: Working with categorical variables"
author: "Your Name Here"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: false
    link-external-newwindow: true
    link-external-icon: true
editor: source
---

## Getting started

```{r}
#| label: setup
#| message: false
#| warning: false

library(ggplot2)
library(dplyr)
library(readr)

crickets <- read_csv("data/crickets.csv")
crickets
```

## The Cricket dataset

### Question 1

Recreate the plot above using a colour scale for the species rather than plotting on separate facets.

```{r}
#| label: plot-colour

crickets |>
  ggplot(aes(x = temp, y = rate, colour = species)) +
  geom_point() +
  scale_x_continuous("Temperature (°C)") +
  scale_y_continuous("Chirp rate") +
  scale_colour_discrete("Species") +
  ggtitle("Cricket chirp rate vs temperature by species") +
  theme_bw()
```

> Colouring by species (rather than faceting) makes it much easier to see how the two species compare.

## Linear modelling

### Question 2

Fit the one-way ANOVA model `rate ~ species`.

```{r}
#| label: anova-model

anova_mod <- lm(rate ~ species, data = crickets)
summary(anova_mod)
```

```text
Call:
lm(formula = rate ~ species, data = crickets)

Residuals:
    Min      1Q  Median      3Q     Max 
-20.486  -9.458  -1.729  13.342  22.271 

Coefficients:
                 Estimate Std. Error t value Pr(>|t|)    
(Intercept)        85.586      3.316  25.807  < 2e-16 ***
speciesO. niveus  -23.156      4.478  -5.171 1.58e-05 ***
---
Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1

Residual standard error: 12.41 on 29 degrees of freedom
Multiple R-squared:  0.4797,    Adjusted R-squared:  0.4617 
F-statistic: 26.74 on 1 and 29 DF,  p-value: 1.579e-05
```

> Note: A one-way ANOVA is a linear regression where the predictor is categorical.

### Question 3

Which level of `species` has been chosen as the reference category? How can you tell from the `lm()` output?

> **`O. exclamationis`** is the reference category. In R, `lm()` defaults to treatment (dummy) coding for factors, which takes the *first level* of the factor as the baseline. Since it is dummy coding:
>
> - `β0` or `(Intercept)` is the mean estimate for the reference group (`O. exclamationis`)
> - `β1` or `speciesO. niveus` is the mean difference between `speciesO. niveus` and the reference group.

### Question 4

Write out the full fitted model equation, defining all terms clearly.

```math
rate^i=85.59−23.16×1[speciesi=O.\ niveus]
```

- `rate^i` is the predicted chirp rate for observation `i`
- `1[⋅]` is an indicator function equal to 1 if cricket `i` is *O. niveus* and 0 if not (i.e., *O. exclamationis* the reference level).

> The mean rate for *O. exclamationis* is `85.59`, and the mean rate for *O. niveus* is `23.16` lower, which is `85.59−23.16=62.43`.

### Question 5

Calculate the estimated mean `rate` for each species using only the summary of the model. Confirm your results with the `predict()` function.

```{r}
#| label: estimated-means

# By hand from the coefficients:
intercept <- coef(anova_mod)["(Intercept)"]
niveus_offset <- coef(anova_mod)["speciesO. niveus"]

exclamationis_mean <- intercept
niveus_mean <- intercept + niveus_offset

c(exclamationis = exclamationis_mean, niveus = niveus_mean)
```

```text
exclamationis.(Intercept)        niveus.(Intercept) 
                 85.58571                  62.42941 
```

```{r}
# Confirm with predict()
predict(anova_mod, newdata = tibble(species = c("O. exclamationis", "O. niveus")))
```

```text
       1        2 
85.58571 62.42941 
```

> Both approaches agree: an estimated mean chirp rate of **85.59** for *O. exclamationis* and **62.43** for *O. niveus*. These are each species’ raw sample mean, since a one-way ANOVA with a single categorical predictor and no other covariates estimates group means directly.

### Question 6

Refit the model without a reference category. What sort of coding is this? Are the results the same?

```{r}
#| label: anova-no-intercept

anova_mod_nointercept <- lm(rate ~ species - 1, data = crickets)
summary(anova_mod_nointercept)
```

```text
Call:
lm(formula = rate ~ species - 1, data = crickets)

Residuals:
    Min      1Q  Median      3Q     Max 
-20.486  -9.458  -1.729  13.342  22.271 

Coefficients:
                        Estimate Std. Error t value Pr(>|t|)    
speciesO. exclamationis   85.586      3.316   25.81   <2e-16 ***
speciesO. niveus          62.429      3.010   20.74   <2e-16 ***
---
Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1

Residual standard error: 12.41 on 29 degrees of freedom
Multiple R-squared:  0.9742,    Adjusted R-squared: 0.9725 
F-statistic: 548.1 on 2 and 29 DF,  p-value: < 2.2e-16
```

> Removing the intercept (by including `- 1`) switches from treatment/dummy coding to **means/cell-means coding**. Comparing the two:
>
> - treatment/dummy coding: intercept (reference) plus an offset(s).
> - means coding: each level of `species` gets its own coefficient, which *is* that group’s mean.
>
> Note that the underlying model (including fitted values, residuals, `R2`) is identical. The only change is the *parameterisation*.
>
> Don’t forget the other coding types mentioned so far: simple coding, deviation coding (sum-to-zero coding), and Helmert coding.

## Adding temperature

### Question 7

Add the temperature term to the regression. Write out the mathematical equation for this model. Is this an additive ANCOVA model?

> With the dummy coding and `O. exclamationis` as reference category. the mathematical model is

```math
ratei=β0+β1×tempi+β2×1[speciesi=O.\ niveus]+ϵi
```

> where `ϵi∼N(0,σ2)` is the error term. In R, we can fit this as

```{r}
#| label: ancova-model

ancova_mod <- lm(rate ~ temp + species, data = crickets)
summary(ancova_mod)
```

```text
Call:
lm(formula = rate ~ temp + species, data = crickets)

Residuals:
    Min      1Q  Median      3Q     Max 
-3.0128 -1.1296 -0.3912  0.9650  3.7800 

Coefficients:
                  Estimate Std. Error t value Pr(>|t|)    
(Intercept)       -7.21091    2.55094  -2.827  0.00858 ** 
temp               3.60275    0.09729  37.032  < 2e-16 ***
speciesO. niveus -10.06529    0.73526 -13.689 6.27e-14 ***
---
Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1

Residual standard error: 1.786 on 28 degrees of freedom
Multiple R-squared:  0.9896,    Adjusted R-squared: 0.9888 
F-statistic: 1331 on 2 and 28 DF,  p-value: < 2.2e-16
```

> The fitted model is

```math
rate^i=−7.21+3.60×tempi−10.07×1[speciesi=O.\ niveus]
```

> Yes, even though we fitted using `lm`, this is an **additive ANCOVA model** with one continuous covariate (`temp`), one categorical factor (`species`), and no interaction between them. Geometrically, it fits two parallel regression lines (one per species) that share a common slope (3.60 chirps/°C) but have different intercepts, offset by a constant 10.07 chirps.

### Question 8

Plot the predictions of the model with an additive effect over the data. Is there evidence that the additive effect should be an interaction? Check by fitting the interaction model.

```{r}
#| label: additive-predictions

crickets |>
  mutate(pred_additive = predict(ancova_mod)) |>
  ggplot(aes(x = temp, y = rate, colour = species)) +
  geom_point() +
  geom_line(aes(y = pred_additive)) +
  scale_x_continuous("Temperature (°C)") +
  scale_y_continuous("Chirp rate") +
  scale_colour_discrete("Species") +
  ggtitle("Additive ANCOVA fit") +
  theme_bw()
```

> Fitting the model with an interaction between `temp` and `species` is done with the `*` term:

```{r}
#| label: interaction-model

interaction_mod <- lm(rate ~ temp * species, data = crickets)
summary(interaction_mod)
```

```text
Call:
lm(formula = rate ~ temp * species, data = crickets)

Residuals:
    Min      1Q  Median      3Q     Max 
-3.7031 -1.3417 -0.1235  0.8100  3.6330 

Coefficients:
                       Estimate Std. Error t value Pr(>|t|)    
(Intercept)           -11.0408     4.1515  -2.659    0.013 *  
temp                     3.7514     0.1601  23.429   <2e-16 ***
speciesO. niveus        -4.3484     4.9617  -0.876    0.389    
temp:speciesO. niveus  -0.2340     0.2009  -1.165    0.254    
---
Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1

Residual standard error: 1.775 on 27 degrees of freedom
Multiple R-squared:  0.9901,    Adjusted R-squared: 0.989 
F-statistic: 898.9 on 3 and 27 DF,  p-value: < 2.2e-16
```

> We can formally compare the models with

```{r}
anova(ancova_mod, interaction_mod)
```

```text
Analysis of Variance Table

Model 1: rate ~ temp + species
Model 2: rate ~ temp * species
  Res.Df    RSS Df Sum of Sq      F Pr(>F)
1     28 89.350                         
2     27 85.074  1    4.2758 1.357 0.2542
```

> The p-value (`0.2542`) of this test indicates that the interaction is not needed. Formally, the null hypothesis `H0:β3=0` (where `β3` is for the interaction effect `temp:speciesO. niveus`) is not rejected at the `α=0.05` level since `0.2542 > 0.05`.
>
> Visually we could already see this. The two **parallel** lines from the additive model already track each species’ points closely, and there’s no obvious systematic pattern that would suggest the slopes genuinely differ.

### Question 9

Why might a linear model be inappropriate for the data?

> Though the linear ANCOVA fits this particular dataset very well (`R2≈0.99`):
>
> - **Chirp rate can’t be negative**, but a linear model will predict negative values of chirp rate for temperatures that are low enough.
>
> Careful use of the model (which is a very good predictor) will avoid any problems with this.

## GLM with factors

```{r}
#| label: poisson-models

model_wool <- glm(breaks ~ wool, data = warpbreaks, family = poisson(link = "log"))
model_tension <- glm(breaks ~ tension, data = warpbreaks, family = poisson(link = "log"))
model_wool_tension_add <- glm(breaks ~ wool + tension, data = warpbreaks, family = poisson(link = "log"))
```

### Question 10

```{r}
# Wool model
summary(model_wool)
```

```text
Call:
glm(formula = breaks ~ wool, family = poisson(link = "log"), 
    data = warpbreaks)

Coefficients:
            Estimate Std. Error z value Pr(>|z|)    
(Intercept)  3.43518    0.03454  99.443  < 2e-16 ***
woolB       -0.20599    0.05157  -3.994 6.49e-05 ***
---
Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1

(Dispersion parameter for poisson family taken to be 1)

    Null deviance: 297.37  on 53  degrees of freedom
Residual deviance: 281.33  on 52  degrees of freedom
AIC: 560

Number of Fisher Scoring iterations: 4
```

> The coefficients in `model_wool` represent
>
> - `(Intercept) = 3.435`, `woolB = -0.206`. On the log scale, wool A’s log-mean breaks is 3.435; wool B’s is 0.206 lower.
> - Exponentiating to get back to counts scale: `exp(3.435)=31.0` mean breaks for wool A, and wool B’s mean is `exp(-0.206)=0.814` times that, i.e. **about 18.6% fewer breaks** for wool B (`31.0×0.814=25.3`).

```{r}
# Tension model
summary(model_tension)
```

```text
Call:
glm(formula = breaks ~ tension, family = poisson(link = "log"), 
    data = warpbreaks)

Coefficients:
            Estimate Std. Error z value Pr(>|z|)    
(Intercept)  3.59426    0.03907  91.988  < 2e-16 ***
tensionM    -0.32132    0.06027  -5.332 9.73e-08 ***
tensionH    -0.51849    0.06396  -8.107 5.21e-16 ***
---
Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1

(Dispersion parameter for poisson family taken to be 1)

    Null deviance: 297.37  on 53  degrees of freedom
Residual deviance: 226.43  on 51  degrees of freedom
AIC: 507.09

Number of Fisher Scoring iterations: 4
```

> The coefficients in `model_tension` represent
>
> - `(Intercept) = 3.076` (Low), `tensionM = -0.321`, `tensionH = -0.518`.
> - High tension has largest difference to low tension: `exp(-0.518)=0.596` times as many breaks as Low tension (a **40% decrease**).

```{r}
# Additive wool + tension model
summary(model_wool_tension_add)
```

```text
Call:
glm(formula = breaks ~ wool + tension, family = poisson(link = "log"), 
    data = warpbreaks)

Coefficients:
            Estimate Std. Error z value Pr(>|z|)    
(Intercept)  3.69196    0.04541  81.302  < 2e-16 ***
woolB       -0.20599    0.05157  -3.994 6.49e-05 ***
tensionM    -0.32132    0.06027  -5.332 9.73e-08 ***
tensionH    -0.51849    0.06396  -8.107 5.21e-16 ***
---
Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1

(Dispersion parameter for poisson family taken to be 1)

    Null deviance: 297.37  on 53  degrees of freedom
Residual deviance: 210.39  on 50  degrees of freedom
AIC: 493.06

Number of Fisher Scoring iterations: 4
```

> The coefficients in `model_wool_tension_add` represent:
>
> - `(Intercept) = 3.173` (`tensionL` and `woolA` reference), `woolB = -0.206`, `tensionM = -0.321`, `tensionH = -0.518`.
> - The `wool` and `tension` coefficients are (as expected for an additive model) essentially unchanged from their single-predictor models. The effects (additive on log scale, multiplicative on raw/count scale) are interpreted *holding the other factors fixed*: at any given tension, wool B has 18.6% fewer breaks than wool A; at any given wool, Medium tension has 27% less breaks than Low whilst Medium has 40% less.

### Question 11

Write out the fitted model equations.

`model_wool`

```math
log(breaks^i)=3.435−0.206×1[wooli=B]
```

`model_tension`

```math
log(breaks^i)=3.594−0.321×1[tensioni=M]−0.519×1[tensioni=H]
```

`model_wool_tension_add`

```math
log(breaks^i)=3.692−0.206×1[wooli=B]−0.321×1[tensioni=M]−0.519×1[tensioni=H]
```

> with reference levels `wool = A` and/or `tension = L` in each case.
>
> These models can also be written on the raw (expected count) scale. For example the last model would be

```math
breaks^i=exp(3.692)×exp(−0.206)1[wooli=B]×exp(−0.321)1[tensioni=M]×exp(−0.519)1[tensioni=H]
```

## References

McDonald, John H. 2014. *Handbook of Biological Statistics*. 3rd ed. Sparky House Publishing.
